Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-08 demystify.tex
updates

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision by around \emph{April 15} (\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

\paragraph{Tenure-line and teaching faculty} Tenure-line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring PhD students.  They often have a low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  At most universities, \textbf{only} tenure and tenured track faculty can serve as a formal adviser of PhD students. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

In contrast, \emph{teaching} (sometimes called instructional faculty or professor of practice) mainly focuses on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. As such they typically do not have funding (e.g., RA~\autoref{sec:funding}) to support PhD students.
Depends on the university, they might be able to \emph{co-advise} and mentor PhD students with a tenure-line faculty.

Also note that non-CS profs unlikely can serve as formal PhD advisers for CS students (see \autoref{sec:related-fields}).

\paragraph{Titles} Faculty have titles such as assistant, associate, full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching faculty or lecturer, also starts with assistant, then promoted to associated, and full prof.

Adjunct faculty is not full-time, e.g., they might be working in industry and teach a class or two for fun.  Emeritus means they are retired but still have some affiliation with the university.   Due to their roles, adjunct, emeritus, and research faculty typically do not advise PhD students.

\emph{Research faculty} or research scientists are typically non-tenure line faculty who focus on research (e.g., publishing papers and getting grants). While they cannot serve as a formal adviser, they often co-advise and mentor students with tenure-line faculty. They can recruit students and provide RA funding for them.

@@

\paragraph{Accepting an offer and later later rejecting it}
I have seen \emph{many} international students asking about this, e.g., accepting an offer from a prof. and then later rejecting it when they get admitted to a better/dream school, which have them on its wait list. There are numerous advice, e.g., on Internet forums, that this is OK to do so (most of these advice are from other international students).
The reasons typically are \emph{``you haven't established much relationship with the prof. yet, so it is OK to reject it; no hard feeling, you probably won't see them again anyway''} or that \emph{``it's your life and you should do what's best for you''}.

From my experience, this is \emph{not} a good idea and can be considered \emph{rude and unprofessional}. But ultimately, it is a trade-off that you need to decide,  e.g., between your own benefit and the cost to others. If you reject an offer, the prof. will likely stop looking for other students and might even reject other students; so by rejecting the offer later, causing a great deal of inconvenience to the prof and also taking away the opportunity from other students. This will likely would burn bridges with that prof and they will remember what you did to them.  Moreover, given that research subfields are typically small, you might run into that professor again in the future, and they will remember what you did to them.

@@

\section{Relationships}
There's a misconception that in the US it's all business, with professors as bosses who pay students for their work and that labmates are just work colleagues; and that doing nice things means expecting something in return.

However, the reality is quite the opposite. While people can be straightforward and appear ``cold'', they are also informal, friendly, and very caring (in ways that might surprise you).
With labmates and colleagues, you will often work and go to lunch together, confide in each other, help each other navigate the academic journey, and often become lifelong friends.
With your professor(s), you can call them by their first name (\autoref{sec:address}), disagree with them and argue (and gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks (\autoref{sec:gifts}).  
Many people maintain lifelong relationships with their professors and colleagues, staying in touch through cards, emails, and calls, even after their academic journey ends. %These connections are genuine and meaningful.

@@

\paragraph{History} This handbook was conceived during a lunch with a faculty at GMU.  We talked about why GMU was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about the US PhD programs (also for US faculty to understand more about international students and therefore have a better chance of recruiting and working with them).

I was also a member of the large VietPhD group on Facebook and browse Internet forums (e.g., Reddit/gradadmission and GradCafe). There I saw many questions from students about PhD programs.  However, most active participants are students in non-CS fields or not in the US, and their answers are unfortunately not always accurate and sometimes lead to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.

I started writing this handbook in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). The document was initially for international students but has been generalized to all students.
I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.

\paragraph{About Me} I am an assistant professor in the CS dept at George Mason University (GMU). Prior to GMU, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at GMU and UNL for many years.  Currently I serve as the program director of the MS program in Software Engineering at GMU (thus also have experience with the MS admission process---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

Though I was not an international student, many of my students and collaborators are/were. I also mentor multiple students from Vietnam and have close colleagues and friends who were once international students. I hope to capture the diverse challenges and experiences they've faced in this document, so that it can be a valuable resource for prospective international students.
Finally, my upbringing in the US provides a perspective aligned with American culture, allowing me to shed light on various issues, particularly those related to cultural differences (\autoref{sec:cultural}).

>>>>

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

\paragraph{Types of Faculty} There are typically three types of faculty: tenure-line, teaching, and research.  The main difference is their focus: research, teaching, or both.  The type of faculty can affect their ability to serve as your adviser.

\textbf{Tenure-line faculty}, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring PhD students.  They often have a low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  For most schools, tenure-line is the main type of faculty that can serve as a formal adviser of PhD students. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

\textbf{Teaching faculty} (sometimes called instructional faculty or professor of practice) mainly focuses on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. As such they typically do not have funding (e.g., RA~\autoref{sec:funding}) to support PhD students.
Depends on the university, they might be able to \emph{co-advise} and mentor PhD students with a tenure-line faculty, but typically do not formally advise and graduate PhD students.

Note that non-CS profs (even in tenure-line ones) often cannot solely advise and graduate PhD students (see \autoref{sec:related-fields}).

Some universities have \textbf{research faculty} or research scientists, who are typically non-tenure line faculty and focus on research (e.g., publishing papers and getting grants). 
They can recruit students and provide RA funding for them. 
They can also teach classes, but typically have a lower teaching load.
For some universities, they \emph{can} solely advise and graduate PhD students, but for others, they cannot.

\textbf{Adjunct faculty} is not full-time, e.g., they might be working in industry and teach a class or two for fun. \textbf{Emeritus faculty} means they are retired from the university, e.g., they could move to the industry, but still have some affiliation with the university.   Due to their roles, these profs. typically do not advise PhD students.

\paragraph{Ranks} Faculty have ranks such as assistant, associate, full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with assistant, then promoted to associated, and full prof.

@@

\paragraph{Accepting an offer and later rejecting it}

I've noticed many students, especially international, face a dilemma when they commit to a graduate offer but then receive a better one, often discussed in online forums. Advice often suggests it's okay to switch, using reasons like not having a strong relationship with the prof. or prioritizing personal benefit.

In my opinion, these reasons are not strong enough to justify retracting an acceptance. A more valid reason is using the \href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{April 15 resolution}, which many universities participate. Among various things, this resolution states that students are free to accept a new offer from a different institution until 4/15, even if they have already accepted an offer elsewhere.

However, regardless of the resolution, retracting an acceptance can have ethical implications. When you accept an offer, you are making a commitment to work with that prof, who then might stop looking for other students. So by retracting your offer, you are breaking your commitment and also causing a great deal of inconvenience to the prof and also takes away the opportunity from other students. 
Ultimately, this choice is personal, involving a balance between personal benefit and the ethical impact on others.

@@

\section{Relationships}
There's a misconception that in the US it's all business, with professors as bosses who pay students for their work and that lab mates are just work colleagues; and that doing nice things means expecting something in return.

However, the reality is quite the opposite. While people can be straightforward and appear ``cold'', they are also informal, friendly, and very caring (in ways that might surprise you).
With lab mates and colleagues, you will often work and go to lunch together, confide in each other, help each other navigate the academic journey, and often become lifelong friends.
With your professor(s), you can call them by their first name (\autoref{sec:address}), disagree with them and argue (and gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks (\autoref{sec:gifts}).  
Many people maintain lifelong relationships with their professors and colleagues, staying in touch through cards, emails, and calls, even after their academic journey ends. %These connections are genuine and meaningful.

@@

\paragraph{History} This document was conceived during a lunch with a faculty at GMU.  We talked about why GMU was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about the US PhD programs (also for US faculty to understand more about international students and therefore have a better chance of recruiting and working with them).

I was also a member of the large VietPhD group on Facebook and browsed Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most active participants are students in non-CS fields or not in the US, and their answers are unfortunately not always accurate and sometimes lead to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). The document was initially for international students but has been generalized to all students.
I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.

\paragraph{About Me} I am an assistant professor in the CS dept at George Mason University (GMU). Before GMU, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at GMU and UNL for many years.  Currently I serve as the program director of the MS program in Software Engineering at GMU (thus also have experience with the MS admission process---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

Though I was not an international student, many of my students and collaborators are/were. I also mentor multiple students from Vietnam and have close colleagues and friends who were once international students. I hope to capture the diverse challenges and experiences they've faced in this document so that it can be a valuable resource for prospective international students.
Finally, my upbringing in the US provides a perspective aligned with American culture, allowing me to shed light on various issues, particularly those related to cultural differences (\autoref{sec:cultural}).
